I/O simulation model with Linux page cache, integration and evaluation in SimGrid framework

### 2020-07-01 paper_io_sim.tex
Some comments in methods
Methods revised

	\section{Method}
		In this section, we present our approach to model file read/write, 
		I/O in memory, data flushing and cache eviction mechanisms 
		implemented in the Linux kernel. 
		We also detail the design the simulators in order to simulate I/O activities 
		and the implementation in Python and SimGrid framework \tristan{check grammar}. 
		Finally, we describe experimental scenarios to evaluate our model using real applications and
		a Python simulator prototype, the current SimGrid simulator, our extension of SimGrid, and the results from real pipelines.

@@

			We model memory with a capacity, a read bandwidth and a write bandwidth. 
			When data is read from or written to memory, the actual 
			read/write time is calculated with the amount of data and memory 
			read/write bandwidth:

			\tristan{how about parallel accesses to memory? I think it would be a motivation to simualte memory as a storage device, 
			and leverage existing bandwidth sharing mechanisms in SimGrid. You could updated Fig 1 to include memory in the storage devices, and 
			explain in your Python simulator implementation that only a simplistic model of storage device is used (for both disk and memory).}

@@

			Next, we simulate data flushing mechanism, which writes back 
			dirty data to disk. 
			The data flushing simulation function traverses the sorted inactive first, 
			and then the active list, looks for least recently used dirty blocks and 
			sets dirty flags of these blocks to 0 until the total flushed amount 
			reaches the amount to be flushed or there is no dirty data left in cache. 
			If the last flushed block is not entirely flushed, it is split into 
			two blocks, one is flushed and one remains dirty.
			The flushing time is calculated as:

			\begin{align*}
				t_f = A / w_d
			\end{align*}

			where:
			\begin{itemize}
				\item $t_f$ is the flushing time
				\item $A$ is the amount of data to be flushed
				\item $w_d$ is the disk write bandwidth
			\end{itemize}

			For periodical flushing, our simulator searches in two page cache 
			LRU lists and flushes all the expired dirty data blocks. 
			A dirty data block is considered expired if the time from its last access 
			to current simulated time is longer than the predefined expire time. 
			The flushing time is calculated by storage simulation model with 
			total amount of flushed data.

@@

			\begin{algorithm}\caption{File read simulation}\label{alg:read}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{fn}{file name}
        				\Desc{fs}{file size (assumed to fit in memory)}
						\Desc{rt}{current simulated run time}
						\Desc{mm}{MemoryManager object}
						\Desc{rm}{memory read bandwidth}
						\Desc{sm}{storage simulation model}
   					\EndInput
   					\Output
						\Desc{rt}{simulated run time after the file is read}
   					\EndOutput
					\State rt = rt + mm.flush(2*fs - mm.cached(fn) - mm.evictable\_mem) 
					\State mm.evict(2*fs - mm.cached(fn) - mm.free\_mem) 
					\If {mm.cached(fn) $>$ 0}  \Comment{Read file cached data}
    					\State mm.cache\_read(fn)  \Comment{Update blocks in LRU lists}
    					\State pft = mm.periodic\_flush(rt) \Comment{Periodical flushing time}
						\State rt = rt + max(mm.cached(fn) / rm, pft) \Comment{Add the longer}
					\EndIf
					\If {mm.cached(fn) $<$ fs} \Comment{Read uncached data from disk}
						\State rt = rt + mm.periodic\_flush(rt) \Comment{Another flushing interval}
						\State mm.disk\_read(fs - mm.cached(fn), fn)
    					\State rt = rt + sm.read(fs - mm.cached(fn))
					\EndIf

@@

			The algorithm used to simulate file read is described in 
			Algorithm~\ref{alg:read}. 
			In Linux systems with write back storage devices, data can be either read 
			from cache with memory bandwidth or read from disk with disk bandwidth. 
			Therefore, total read time can be considered as a linear function 
			of the read time with each of these bandwidths, 
			enabling us to generalize reading process as two separate phases, 
			cache read and disk read. 
			When a user simulated application reads a file, 
			it sends a file read request to IOController. 
			First, dirty data is flushed to disk to make more available memory 
			to read the file if needed.
			Data flushing function in MemoryManager is called, flushing time is 
			simulated and added to total run time (line 10).
			After data is flushed, MemoryManager is asked to evict data 
			from cache to accommodate the requested file (line 11). 
			Next, if file is cached either partially or entirely, MemoryManager is 
			called to re-access and update this cached data (line 13). 
			Then, periodical flushing is called to flushed expired dirty data 
			from cache (line 14).
			Since periodical flushing and cache read can be concurrent, the time of 
			the longer activity between them is added to simulated run time (line 15). 
			If file is not cached or partially cached, a disk read is required (line 17). 
			Periodical flushing is called once again, but this time the flushing time is 
			added to total simulated time since flushing and disk read are time-shared, 
			or sequential in other words (line 18).
			Finally, data is read from disk to cache and managed by MemoryManager, 
			disk read time estimated by the storage model is added to 
			total simulated run time (line 19-20).

			\begin{algorithm}\caption{File write simulation}\label{alg:write}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{fn}{file name}
        				\Desc{fs}{file size}
						\Desc{rt}{current simulated run time}
						\Desc{mm}{MemoryManager object}
						\Desc{wm}{memory write bandwidth}
						\Desc{sm}{storage simulation model}
   					\EndInput
   					\Output
						\Desc{rt}{simulated run time after the file is read}
   					\EndOutput
   					\State pft = mm.periodic\_flush(rt)
					\State remain\_dirty = dirty\_ratio * mm.avail\_mem - mm.dirty
					\State mem\_amt = min(fs, remain\_dirty)
					\If {remain\_dirty $>$ 0} \Comment{Write with memory bandwidth}
    					\State mm.evict(mem\_amt - mm.free\_mem)
    					\State mm.write(fn, mem\_amt)
    				\EndIf
    				\State rt = rt + max(mem\_amt / wm, pft) \Comment{Concurrent processes}
					\If {mem\_amt $<$ fs}  \Comment{Write with disk bandwidth}
						\State mm.flushed(fs - mem\_amt)  \Comment{Concurrent flushing}
						\State mm.evict(fs - mem\_amt  - mm.free\_mem) 
						\State mm.write(fn, min(fs - mem\_amt, mm.free\_mem))
						\State rt = rt + sm.write(fs - mem\_amt)
					\EndIf

@@

			Algorithm~\ref{alg:write} describes our simulation of file write. 
			Initially, periodical flushing is called to flush expired dirty data 
			at the beginning of the write (line 10).
			Similar to the file read simulation, a file can also be written with 
			two different bandwidths, memory bandwidth and disk bandwidth. 
			Files can be written with memory bandwidth before the amount 
			of dirty data reaches dirty\_ratio. Thus, our algorithm initially checks 
			the amount of file data that can be written with memory write 
			bandwidth (line 11-12).
			If this amount is greater than 0, MemoryManager is asked to evict 
			data from cache if needed and to write this amount of data to cache 
			(line 14-15).
			The simulated run time can be added with the longer time of cache write 
			time and periodical flushing time because these can happen 
			concurrently (line 17).
			The amount of data that is not written with memory bandwidth is written 
			with disk bandwidth (line 18). 
			Now, the dirty ratio threshold is reached, we have to flush and evict 
			data from cache as much as possible so that 
			the remaining data can be written to cache (line 19-20). 
			Here, MemoryManager does not guarantee that the amount of flushed 
			and evicted data is enough for the remaining data. In the case it is 
			less than the required amount, the remaining data is written to cache 
			but then can be flushed and evicted right away. Thus, the amount 
			of remaining data that is written and kept in cache after the write 
			is limited by the amount free memory (line 21). 
			Finally, the write time is simulated, added to the total time and 
			returned to simulated applications (line 22).

@@

			To validate our simulation model, we create a simple 
			simulator which is independent to simulation frameworks and libraries. 
			This enables us to evaluate the accuracy and correctness of our 
			model in a simple use case before integrating it to SimGrid. 
			This simulator simulates a single-threaded pipeline running on a system 
			with a single core CPU, one local storage device and all input and 
			output files are stored locally. The results from the simulator is compared 
			and validated with the results from a real pipeline running on
			a real system.

>>>>

	\section{Method}
		In this section, we present our approach to model file read/write, 
		I/O in memory, data flushing and cache eviction mechanisms 
		implemented in the Linux kernel. 
		We also detail the design of the simulators in order to simulate 
		I/O activities and the implementation in Python and SimGrid framework. 
		Finally, we describe experimental scenarios to evaluate our model using 
		real applications and a Python simulator prototype, the current SimGrid 
		simulator, our extension of SimGrid, and the results from real pipelines.

@@

			Memory can be characterized with a capacity, read bandwidth and write 
			bandwidth. However, memory bandwidth is always shared between 
			concurrent processes and implementing a simulation model of concurrent 
			accesses to memory is non-trivial. Thus, we leverage existing storage 
			models, which handle parallel storage accesses, in some simulation 
			frameworks to simulate I/O time in memory.

@@

			Next, we simulate data flushing mechanism, which writes back 
			dirty data to disk. 
			The data flushing simulation function traverses the sorted inactive first, 
			and then the active list, looks for least recently used dirty blocks and 
			sets dirty flags of these blocks to 0 until the total flushed amount 
			reaches the amount to be flushed or there is no dirty data left in cache. 
			If the last flushed block is not entirely flushed, it is split into 
			two blocks, one is flushed and one remains dirty.
			The flushing time is simulated with storage model.

			For periodical flushing, our simulator searches in two page cache 
			LRU lists and flushes all the expired dirty data blocks. 
			A dirty data block is considered expired if the time from its last access 
			to current simulated time is longer than the predefined expire time. 
			Periodical flushing can be concurrent with cache read/write, 
			and sequential with disk read/write. When cache accesses and 
			periodical flushing are concurrent, flushing time is not added to simulated 
			time but the amount of flushed data is limited by flushing time. 
			When data is flushed in parallel with disk read/write, flushing time is 
			calculated by storage simulation model and total amount of data 
			to be flushed.

@@

			\begin{algorithm}\caption{File read simulation}\label{alg:read}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{fn}{file name}
        				\Desc{fs}{file size (assumed to fit in memory)}
						\Desc{rt}{current simulated run time}
						\Desc{mm}{MemoryManager object}
						\Desc{rm}{memory read bandwidth}
						\Desc{sm}{storage simulation model}
   					\EndInput
   					\Output
						\Desc{rt}{simulated run time after the file is read}
   					\EndOutput
					\State rt = rt + mm.flush(2*fs - mm.cached(fn) - mm.evictable\_mem) 
					\State mm.evict(2*fs - mm.cached(fn) - mm.free\_mem) 
					\If {mm.cached(fn) $>$ 0}  \Comment{Read file cached data}
    					\State crt = mm.cache\_read(fn)  \Comment{Read data in cache}
    					\State mm.periodic\_flush(rt, crt) \Comment{Flush in duration = crt}
						\State rt = rt + crt
					\EndIf
					\If {mm.cached(fn) $<$ fs} \Comment{Read uncached data from disk}
						\State rt = rt + mm.periodic\_flush(rt) \Comment{Sequential flushing}
						\State mm.disk\_read(fs - mm.cached(fn), fn)
    					\State rt = rt + sm.read(fs - mm.cached(fn))
					\EndIf

@@

			The algorithm used to simulate file read is described in 
			Algorithm~\ref{alg:read}. 
			In Linux systems with write back storage devices, data can be either read 
			from cache with memory bandwidth or read from disk with disk bandwidth. 
			Therefore, total read time can be considered as a linear function 
			of the read time with each of these bandwidths, 
			enabling us to generalize reading process as two separate phases, 
			cache read and disk read. 
			When a user simulated application reads a file, 
			it sends a file read request to IOController. 
			First, dirty data is flushed to disk to make more available memory 
			to read the file if needed.
			Data flushing function in MemoryManager is called, flushing time is 
			simulated and added to total run time (line 10).
			After data is flushed, MemoryManager is asked to evict data 
			from cache to accommodate the requested file (line 11). 
			Next, if file is cached either partially or entirely, MemoryManager is 
			called to re-access and update this cached data (line 13). 
			During cache read, periodical flushing is called to flush dirty data if needed 
			and limited to cache read time (line 14).
			Cache read time is then added to total simulated time (line 15). 
			If file is not cached or partially cached, a disk read is required (line 17). 
			Periodical flushing is called once again, but not time-limited since 
			data flushing and disk read are sequential, and the flushing time is 
			added to simulated time (line 18).
			Finally, data is read from disk to cache and managed by MemoryManager, 
			disk read time estimated by the storage model is added to 
			total simulated run time (line 19-20).

			\begin{algorithm}\caption{File write simulation}\label{alg:write}
				\small
				\begin{algorithmic}[1]
					\Input
        				\Desc{fn}{file name}
        				\Desc{fs}{file size}
						\Desc{rt}{current simulated run time}
						\Desc{mm}{MemoryManager object}
						\Desc{wm}{memory write bandwidth}
						\Desc{sm}{storage simulation model}
   					\EndInput
   					\Output
						\Desc{rt}{simulated run time after the file is read}
   					\EndOutput
   					\State pft = mm.periodic\_flush(rt)
					\State remain\_dirty = dirty\_ratio * mm.avail\_mem - mm.dirty
					\State mem\_amt = min(fs, remain\_dirty)
					\If {remain\_dirty $>$ 0} \Comment{Write with memory bandwidth}
    					\State mm.evict(mem\_amt - mm.free\_mem)
    					\State mm.periodic\_flush(rt, mm.write\_time(mem\_amt)) 
    					\State rt = rt + mm.write(fn, mem\_amt)
    				\EndIf
    				\State rt = rt + max(mem\_amt / wm, pft) \Comment{Concurrent processes}
					\If {mem\_amt $<$ fs}  \Comment{Write with disk bandwidth}
						\State mm.flushed(fs - mem\_amt)  \Comment{Concurrent flushing}
						\State mm.evict(fs - mem\_amt  - mm.free\_mem) 
						\State mm.write(fn, min(fs - mem\_amt, mm.free\_mem))
						\State rt = rt + sm.write(fs - mem\_amt)
					\EndIf

@@

			Algorithm~\ref{alg:write} describes our simulation of file write. 
			Initially, periodical flushing is called to flush expired dirty data 
			at the beginning of the write (line 10).
			Similar to the file read simulation, a file can also be written with 
			two different bandwidths, memory bandwidth and disk bandwidth. 
			Files can be written with memory bandwidth before the amount 
			of dirty data reaches dirty\_ratio. Thus, our algorithm initially checks 
			the amount of file data that can be written with memory write 
			bandwidth (line 11-12).
			If this amount is greater than 0, MemoryManager is asked to evict 
			data from cache if needed (line 14).
			Next, periodical flushing flushes data during cache write because they 
			can be concurrent (line 15). Then, data is written to cache with memory 
			bandwidth, write time is simulated and added to total simulated 
			time (line 16).
			The amount of data that is not written with memory bandwidth is written 
			with disk bandwidth (line 18). 
			Now, the dirty ratio threshold is reached, we have to flush and evict 
			data from cache as much as possible so that 
			the remaining data can be written to cache (line 20-21). 
			Here, MemoryManager does not guarantee that the amount of flushed 
			and evicted data is enough for the remaining data. In the case it is 
			less than the required amount, the remaining data is written to cache 
			but then can be flushed and evicted right away. Thus, the amount 
			of remaining data that is written and kept in cache after the write 
			is limited by the amount free memory (line 22). 
			Finally, the write time is simulated, added to the total time and 
			returned to simulated applications (line 23).

@@

			To validate our simulation model, we create a simple 
			simulator which is independent to simulation frameworks and libraries. 
			This enables us to evaluate the accuracy and correctness of our 
			model in a simple use case before integrating it to SimGrid. 
			Since the simulator does not use any storage model, we simulate the 
			I/O time in memory as follow:

			This simulator simulates a single-threaded pipeline running on a system 
			with a single core CPU, one local storage device and all input and 
			output files are stored locally. The results from the simulator is compared 
			and validated with the results from a real pipeline running on
			a real system.
